---
title: "Final Project"
author: "Madina Kanafina"
date: "`r Sys.Date()`"
documentclass: article
geometry: margin=1in
fontsize: 11pt
output:
  pdf_document:
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    dev: pdf
    highlight: tango
  html_document:
    theme: default
    self_contained: true
    toc: false
    df_print: kable
    fig_caption: false
    number_sections: false
    smart: true
    dev: svg
---

```{r setup, include = FALSE}
# DO NOT ALTER THIS CHUNK
# Set knitr options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  fig.width = 5,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 120,
  fig.align = "center",
  cache = FALSE
)
# Load required packages
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(infer))
suppressPackageStartupMessages(library(modelr))
suppressPackageStartupMessages(library(broom))
# Load dataset
college <- read_rds("college.rds") %>%
  type_convert(
    na=vctrs::vec_c("", "NA", "PrivacySuppressed")
  )
# Set seed
set.seed(98261936)
```

## Introduction

The U.S. Department of Education's College Scoreboard dataset contains a lot of information about colleges across the United States. For this final project, I will investigate the question:

> "Is there a relationship between how much it costs to attend the college and how much its graduates earn 10 years after enrolling?"

This question is relevant and interesting because many students and families want to know if paying more for college leads to better financial outcomes. To answer this question, I will use the variable 'COSTT4_A'(average cost of attendance) as the predictor and 'MD_EARN_WNE_P10'(median earning of graduates after 10 years) as the outcome. I will create a linear regression model to see if there is a significant relationship between these two continuous variables. The results of this analysis could help students decide if higher college costs are worth it.

## Preprocessing

```{r}
college_reduced <- college %>%
  select(COSTT4_A, MD_EARN_WNE_P10) %>%
rename(
  avg_cost = COSTT4_A,
  median_earn_10yr = MD_EARN_WNE_P10
)
```

I've selected only the columns 'COSTT4_A' and 'MD_EARN_WNE_P10' because they are directly relevant to my question, and I've renamed them to make them more human-readible and easier to interpret during the analysis.

## Visualization




## Summary Statistics




## Data Analysis




## Conclusion





